Typed Interactions for NLU in Opal

### 2017-12-12 paper.tex
Added questions document, more additions to the paper

% TODO: acmart, acmsmall

@@

\begin{document}
\maketitle
\begin{abstract}
  This is an abstract containing relatively abstract things.
\end{abstract}

@@

% \subsection{Opal}
% Opal is a research programming language being developed at Cornell University
% under the guidance of Professor Adrian Sampson. The language is being developed
% as an embedded language within Typescript, and it is designed with machine
% learning applications in mind. Opal supports programming paradigms that fit well
% with the interactive and unpredictable nature of ML programs, and has libraries
% for interacting with various types of ML engines.

@@

\subsection{The DSL}
With our observations about search strategies in mind, we can write our types
from before in a slightly different way. Here are the same type declarations
from section \ref{introduction}, but written in our DSL:
\begin{minted}{typescript}
  free-text Msg;
  keywords Time = "later" | "tomorrow";
  alias Set = {
    msg: Msg,
    time: Time
  };
  trait Intent =
    | <Set> Set
    | <Get> {};
\end{minted}

@@

Unfortunately, there is still one major issue. The vast majority of NLU engines
don't return structured data. Instead, they return a bag of entities with no
information about the relationships between them. Figure \ref{fig:nostructure}
shows a sample response. Recall that in our example, \ts{Msg} and \ts{Time}
entities were only expected when there was an \ts{Intent} with tag \ts{"Set"}.
Furthermore, our types dictate exactly where in the data structure those entity
values should end up. The popular libraries, Wit incuded, igore this structure
entirely.

@@

We will define parse by case analysis on the declaration.
\begin{align*}
  \text{parse}(\text{FreeText}(\etag{t}), r) = \text{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a string, we can simply look up the value of the appropriate
entity from the response.
\begin{align*}
  \text{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \text{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$, but since the NLU engine has
been configured with the same list of literals, we are guaranteed to get one of
them back from lookup (assuming lookup is defined).
\begin{align*}
  \text{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \text{parseTy}(rec[\text{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for parse simply matches the value in the response to an index
into the record, and interprets that record entry using parseTy.

@@

\subsection{Correctness}
Given a declaration $d$, we generate a configuration, cfg. Then,
\begin{flalign*}
  &\forall\ (r : \textsf{Resp}).\\
  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
\end{flalign*}

@@

\subsection{Runtime Library}

\section{Evaluation} \label{evaluation}
Talk about how great we did.

\section{Future Work} \label{future}
{\bf Brain Dump}
\begin{itemize}
\item More types (include Wit builtins)
\item More backends
\item Use confidence information
\item Unit tests as training samples
\end{itemize}

\section{Conclusion} \label{conclusion}

>>>>

@@

\begin{document}
\maketitle

\begin{abstract}
  Natural Language Understanding (NLU) is a powerful tool for modern
  applications. Interacting with remote NLU engines can be error prone: most
  backends provide no type guarantees, and configuration synchronization is a
  constant source of bugs. We present a DSL for configuring a NLU application
  that ensures synchronization and type-safety.
\end{abstract}

@@

@@

\subsection{The DSL}
With our observations about search strategies in mind, we can write our types
from before in a slightly different way. Figure \ref{fig:dsl_decls} same type
declarations from section \ref{introduction} written in our DSL.

\begin{figure}
\begin{minted}{typescript}
  free-text Msg;
  keywords Time = "later" | "tomorrow";
  alias Set = {
    msg: Msg,
    time: Time
  };
  trait Intent =
    | <Set> Set
    | <Get> {};
\end{minted}
  \caption{Some DSL declarations.}
  \label{fig:dsl_decls}
\end{figure}

@@

Unfortunately, there is still one major issue. The vast majority of NLU engines
do not return structured data. Instead, they return a bag of entities with no
information about the relationships between them. Figure \ref{fig:nostructure}
shows a sample response. Recall that in our example, \ts{Msg} and \ts{Time}
entities were only expected when there was an \ts{Intent} with tag \ts{"Set"}.
Furthermore, our types dictate exactly where in the data structure those entity
values should end up. The popular libraries, Wit incuded, igore this structure
entirely.

@@

We will define parse by case analysis on the declaration.
\begin{align*}
  \text{parse}(\text{FreeText}(\etag{t}), r) = \text{lookup}(\etag{t}, r)
\end{align*}
This case is fairly straightforward. Since the interpretation of a free-text
entity is just a string, we can simply look up the value of the appropriate
entity from the response.
\begin{align*}
  \text{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
  \text{lookup}(\etag{t}, r)
\end{align*}
Perhaps surprisingly, we can use the exact same definition in the Keywords case
that we did in the FreeText case. We are required to return a value of type
$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
well-formed, it will have been configured to have the same list of literals. We
are guaranteed to get one of them back from lookup.
\begin{align*}
  \text{parse}(\text{Trait}(\etag{t}, rec), r) = \\
  \text{parseTy}(rec[\text{lookup}(\etag{t}, r)], r)
\end{align*}
The final case for parse simply matches the value in the response to an index
into the record, and interprets that record entry using parseTy.

@@

\subsection{Correctness}
Given a declaration $d$, we generate a configuration, cfg. Then,
\begin{flalign*}
  &\forall\ (r : \textsf{Resp}).\\
  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
\end{flalign*}
Note that since parse is a dependent function, it will already return a value of
the correct type. Thus, we only need to care that it is defined. Proof of this
fact is clear by induction on $d$.

@@

\subsection{Runtime Library}
At run-time typical application using NLU will go through a fairly consistent
cycle.
\begin{enumerate}
\item Ask user for input, converting voice to text if necessary.
\item Query the NLU engine and obtain a response for the given input.
\item Parse the response into application data.
\item Process parsed data, and repeat.
\end{enumerate}
Steps 1 and 4 are generally application-specific, so we have elected to provide
a library which handles steps 2 and 4. For each backend that Opal
supports,\footnote{At the time of writing, Wit is the only supported backend.}
we provide utilities for querying the API and parsing the response data. The
network code involved in obtaining the response is fairly uninteresting, so we
will focus on the parsing,

In section \ref{formalism}, we outline the parse function, which takes a
declaration and a response and generates an Opal object. The function that we
provide in the Opal library is similar, but there are two important differences.
First, the parse function takes some element of $\fcy{D}$, which it uses to
deduce the structure of the desired output. In the real application, the analog
of $\fcy{D}$ is the \hs{Decl} type in the Haskell code; this is not available to
the Opal application at run time. As mentioned above, we get around this by
using a modified version the runtypes library to keep this information around at
run-time. We extend runtypes to include a type for entities---which map to
declarations in $\fcy{D}$. Runtypes provides the infrastructure for non-entity
types in $\fcy{T}$.

The other difference arises because Opal does not have dependent types. We
cannot completely faithfully transcribe the parse function from above, because
it is not obvious that it can be typed at all. The solution here is to combine
Opal's dynamic typing with runtypes' dynamic checking feature to create a
function that appears to be dependently typed. First, we use dynamic types
within the parse function. While building the response data, we assume, where
necessary, that it has type \ts{any}. Then, once parsing is complete, we use the
runtypes object to check the parsed object, and we cast it to the desired type.
The definition is shown in figure \ref{fig:parse}. The cast is safe, so long as
the user respects the invariant that parse is always called with compatible type
and runtype parameters.

\begin{figure*}
\begin{minted}{typescript}
          public parse<T>(rt: Runtype<any>, res: Response): T {
              return rt.check(this._parse(rt, res)) as T;
          }

          private _parse(rt: Runtype<any>, res: Response): any { ... }
\end{minted}
  \caption{The wit parse functions.}
  \label{fig:parse}
\end{figure*}

\section{Evaluation} \label{evaluation}
%TODO: Cover limitations of the design, possible use-cases, etc

\section{Future Work} \label{future}
Moving forward, we hope to take this work in two main directions---wider support
for NLU features and backends and entirely new features.

\subsection{Wider Support}
Expanding support for more entity types and backends is an obvious way to
increase the usefulness of our work. Here, we discuss both of those extensions
in detail.

\subsubsection{More Types}
In addition to the three ``universal'' kinds of entities (free-text, keywords,
and traits), the Wit NLU engine provides more specialized ways of looking up
entities. Wit provides support for parsing numbers, dates, times, and other
specific forms of text. We elected to keep things simple at first and exclude
these built-in entities, but adding support for them seems like a very
reasonable next-step.

At a first approximation, this would require modifying the definitions of
$\fcy{D}$ and $\tau$ to include declarations for built-in entities and
base types for the data associated with those entities. More work needs to be
done exploring this space before any more specific conclusions can be drawn.

\subsubsection{More Backends}
We are also interested exploring NLU engines beyond Wit. One reason for this is
purely practical---users might want to use other backends, and it would be nice
for Opal to support this. Another reason, however, is more academic. Exploring
other NLU engines would give us opportunities to test the generality of our
current design. Preliminary research suggests that building our type system
around entities and search strategies makes sense, but we can only know that for
sure if we try to build interfaces for more engines.

\subsection{Interesting Features}
There are a number of novel research directions available as extensions to this
project. The following are two of the many ideas that we have had.

\subsubsection{Confidence-Based Parsing}
Up to this point, we have assumed very little about the data returned in the NLU
response. In practice, however, many NLU engines provide more than just
information about the matched entities. Wit, in particular, provides confidence
levels for each entitiy that it finds. Thus, a value like \ts{"Set"} might be
paired with a confidence level of $70\%$, meaning that Wit is $70\%$ sure that
the user's intent is actually to set a new reminder.

As of now, we do not actually use this information. In the majority of cases
that we have tested, a Wit bot either has enough information to get the right
entities, or it does not. However, it is possible that as we attempt to support
new types of entities and as user types get more complicated, it will be useful
to try multiple parses of a single response. In that case, we can choose between
different parses using confidence scores.

\subsubsection{Training with Tests}
A final extension of this project would attempt to close the code/configuration
gap even further by allowing users to train their NLU application in a way that
is guaranteed to support their tests. We would accomplish this using the same
idea behind our type DSL---we introduce a new DSL (or an extension to the
current one) for generating both Wit training examples and application tests.
While synchronization is less of an issue with training than it is with
configuration, it would still be extremely convenient to express the entirety of
the application's behavior in one place.

Exploring this path would require a fair amount of implementation work, but we
would also need to take great care to avoid overfitting. It is unclear to what
extent this would become an issue.

\section{Conclusion} \label{conclusion}
% TODO: Write
